---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>",
  fig.path = "README-"  
)
```

# janitor
***********************
[![Travis-CI Build Status](https://travis-ci.org/sfirke/janitor.svg?branch=master)](https://travis-ci.org/sfirke/janitor)
[![Coverage Status](https://img.shields.io/codecov/c/github/sfirke/janitor/master.svg)](https://codecov.io/github/sfirke/janitor?branch=master)


Some packages have complex or powerful functions.  janitor has simple little functions to make data cleaning easier.

<hr/>
>Data scientists, according to interviews and expert estimates, spend from 50 percent to 80 percent of their time mired in this more mundane labor of collecting and preparing unruly digital data, before it can be explored for useful nuggets.
>
> -- *"[For Big-Data Scientists, 'Janitor Work' Is Key Hurdle to Insight](http://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html)" - The New York Times, 2014*

<hr/>

## Installation

janitor is not yet on CRAN.  Install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
install_github("sfirke/janitor")
```


# Using janitor

Get started with the [janitor vignette](https://github.com/sfirke/janitor/blob/master/vignettes/introduction.md), which shows how to use the janitor functions.

## Contact

You are welcome to:

* submit suggestions and bug-reports at: https://github.com/sfirke/janitor/issues
* send a pull request on: https://github.com/sfirke/janitor/
* compose a friendly e-mail to: <img src = "http://samfirke.com/wp-content/uploads/2016/07/email_address_as_png.png" alt = "samuel.firke AT gmail" width = "210"/>
