---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>",
  fig.path = "README-"  
)
```

# janitor

janitor helps fix common dirty data issues.  Functions are human-readable - think `clean_names()` instead of re-using code like`gsub("[.]+", "_", .) ...`.

Use with the `%>%` pipe from [magrittr](https://github.com/smbache/magrittr) immediately after reading in data for most elegant results.


```{r, message = FALSE, eval = FALSE}
# load demo packages using the pacman package
if (!require("pacman")) install.packages("pacman"); library(pacman)
p_load(janitor, dplyr, readr)

# read sample dirty data file
starting_df <- read_csv("http://www.github.com/sfirke/janitor/sample/dirty_data.csv")

# the initial names are malformed.  Duplicate names will cause dplyr calls to fail.
names(starting_df)

# clean the data with janitor:
clean_df <- starting_df %>%
  clean_names %>%
  remove_empty_rows %>%
  remove_empty_cols

# names are now clean:
names(clean_df)

```

## Overview

The janitor functions are:

* Clean data.frame names with `clean_names()`.

* Remove entirely empty rows with `remove_empty_rows()` and empty columns with `remove_empty_cols()`.

## Installation

janitor isn't on CRAN yet.  Install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
install_github("sfirke/janitor")
```
