---
title: "Catalog of janitor functions"
date: "`r Sys.Date()`"
output:
  rmarkdown::github_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(janitor)
```
The janitor functions expedite the initial data exploration and cleaning that comes with any new data set.  This catalog describes the usage for each function.

# Major functions
Functions for everyday use.

## Cleaning

### Clean data.frame names with `clean_names()`
Call this function every time you read data.

It works in a `%>%` pipeline, and handles problematic variable names, especially those that are so well preserved by `readxl::read_excel()` and `readr::read_csv()`.

+ Returns names with only lowercase letters, with `_` as a separator
+ Handles special characters and spaces
+ Appends numbers to duplicated names
+ Converts "%" to "percent" to retain meaning

```{r, message = FALSE, warning = FALSE}
# Load dplyr for the %>% pipe
library(dplyr)
# Create a data.frame with dirty names
test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("hIgHlo", "REPEAT VALUE", "REPEAT VALUE",
                    "% successful (2009)",  "abc@!*", "")
```
Clean the variable names, returning a data.frame:
```{r}
test_df %>%
  clean_names()
```
Compare to what base R produces:
```{r}
make.names(names(test_df))
```

## Exploring

### `tabyl()` - a better version of `table()`
`tabyl()` takes a vector and returns a frequency table, like `table()`. But its additional features are:

+ It returns a data.frame - for manipulating further, or printing with `knitr::kable()`.
+ It automatically calculates percentages
+ It can (optionally) display `NA` values
    + When `NA` values are present, it will calculate an additional column `valid_percent` in the style of SPSS
+ It can (optionally) sort on counts
+ It can be called with `%>%` in a pipeline
+ When called on a factor, it will include missing levels in the result (levels not present in the vector)

Usage:
```{r}
x <- c("a", "b", "c", "c", NA)
tabyl(x)
```
Compare to:
```{r}
table(x)
```

`tabyl()` can be called on a piped-in data.frame, which allows for fast, flexible exploration of data:
```{r}
mtcars %>%
  filter(gear > 3) %>%
  tabyl(cyl)
```


### Crosstabulate two variables with `crosstab()`
`crosstab()` generates a crosstab table.  There many R crosstab functions already; this one is distinguished by:

+ It returns a data.frame
+ It is simple.
    + It can (optionally) display `NA` values
+ It can be called with `%>%` in a pipeline

And it has a host of functions to format the result *(more on that later).*

Usage:
```{r}
dat <- data.frame(
  x = c("a", "a", "b", "b", NA),
  y = c(1, 1, 2, 1, 2)
)

dat %>% crosstab(x, y)

# With the most basic of its adornment functions:
dat %>%
  crosstab(x, y) %>%
  adorn_percentages("row")
```

This function wraps the common pipeline of `group_by %>% summarise %>% mutate %>% spread` from the dplyr and tidyr packages, often used in exploratory analysis.  The simple `crosstab` call above produces the same result* as this much longer pipeline:
```{r, message=FALSE, results = "hide"}
library(dplyr) ; library(tidyr)
dat %>%
  group_by(x, y) %>%
  tally() %>%
  mutate(percent = n / sum(n, na.rm = TRUE)) %>%
  select(-n) %>%
  spread(y, percent, fill = 0) %>%
  ungroup()
```
And is more featured than the base R equivalents `table(dat$x, dat$y)` and `prop.table(table(dat$x, dat$y), 1)`.

\**not exactly: the long pipeline returns a `tibble`, while crosstab() returns a `data.frame` that prints fully in the console.*

### Format a crosstab table with `adorn_*` functions

These small, modular functions build on a `crosstab()` to  approximate the functionality of a quick Microsoft Excel PivotTable.  They print an elegant result, either for interactive analysis or for sharing in a report, e.g., with `knitr::kable()`.  For example:

```{r}
mtcars %>%
  crosstab(cyl, gear) %>%
  adorn_totals("row") %>%
  adorn_percentages("row") %>% 
  adorn_pct_formatting() %>%
  adorn_ns()

```

The display options can be fine-tuned:

+ Add totals row, column, or both
+ Calculate percentages along either axis
+ Control how many digits to display and whether to append a % sign
+ Round percentages either with the default `round()` function, or round-half-to-up using a [custom rounding function](http://stackoverflow.com/a/12688836/4470365)
    + e.g., round 10.5 up to 11, consistent with Excel's tie-breaking behavior
    + This contrasts with rounding 10.5 down to 10 as in base R's `round(10.5)`.
    
You can also call these functions on other data.frames, not just the result of calls to `crosstab()`.  E.g., `mtcars %>% adorn_totals(col) %>% adorn_percentages("col")`.

For more on these functions, including how they play together, see **THIS VIGNETTE - LINK TO COME.**

### Explore records with duplicated values for specific combinations of variables with `get_dupes()`
This is for hunting down and examining duplicate records during data cleaning - usually when there shouldn't be any.

For example, in a tidy data frame you might expect to have a unique ID repeated for each year, and year repeated for each unique ID, but no duplicated pairs of unique ID & year.  Say you want to check for their presence, and study any such duplicated records.

`get_dupes()` returns the records (and inserts a count of duplicates) so you can sleuth out the problematic cases:
```{r}
get_dupes(mtcars, wt, cyl) # or mtcars %>% get_dupes(wt, cyl) if you prefer to pipe
```



# Minor functions
Smaller functions for use in particular situations.  More human-readable than the equivalent code they replace.

## Cleaning

### Fix dates stored as serial numbers with `excel_numeric_to_date()`
Ever load data from Excel and see a value like `42223` where a date should be?  This function converts those serial numbers to class `Date`, and contains an option for specifying the alternate date system for files created with Excel for Mac 2008 and earlier versions (which count from a different starting point).
```{r}
excel_numeric_to_date(41103)
excel_numeric_to_date(41103, date_system = "mac pre-2011")
```

### `remove_empty_cols()` and `remove_empty_rows()`
One-line wrapper functions that do what they say.  For cases like cleaning Excel files containing empty rows and columns.
```{r}
q <- data.frame(v1 = c(1, NA, 3),
                v2 = c(NA, NA, NA),
                v3 = c("a", NA, "b"))
q %>%
  remove_empty_cols() %>%
  remove_empty_rows()
```

## Exploring

### `adorn_totals()`
This adds a totals row and/or column to a data.frame (you specify which).  This function excludes the first column of the input data.frame, assuming that it contains a descriptive variable not to be summed.
```{r}
mtcars %>%
  crosstab(am, cyl) %>%
  adorn_totals("row")
```


### Count factor levels in groups of high, medium, and low with `top_levels()`

Originally designed for use with Likert survey data stored as factors.  Returns a `tbl_df` frequency table with appropriately-named rows, grouped into head/middle/tail groups.

+ Takes a user-specified size for the head/tail groups
+ Automatically calculates a percent column
+ Supports sorting
+ Can show or hide `NA` values.

```{r}
f <- factor(c("strongly agree", "agree", "neutral", "neutral", "disagree", "strongly agree"),
            levels = c("strongly agree", "agree", "neutral", "disagree", "strongly disagree"))
top_levels(f)
top_levels(f, n = 1, sort = TRUE)
```

