---
title: "Overview of janitor functions"
date: "`r Sys.Date()`"
output:
  rmarkdown::github_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{janitor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(janitor)
```
The janitor functions expedite the initial data exploration and cleaning that comes with any new data set.  This catalog describes the usage for each function.

# Major functions
Functions for everyday use.

## Cleaning

### Clean data.frame names with `clean_names()`
Call this function every time you read data.

It works in a `%>%` pipeline, and handles problematic variable names, especially those that are so well-preserved by `readxl::read_excel()` and `readr::read_csv()`.

+ Parses letter cases and separators to a consistent format.
    + Default is to snake_case, but other cases like camelCase are available
+ Handles special characters and spaces, including transliterating characters like `œ` to `oe`.
+ Appends numbers to duplicated names
+ Converts "%" to "percent" and "#" to "number" to retain meaning
+ Spacing (or lack thereof) around numbers is preserved

```{r, message = FALSE, warning = FALSE}
# Create a data.frame with dirty names
test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")
```
Clean the variable names, returning a data.frame:
```{r}
test_df %>%
  clean_names()
```
Compare to what base R produces:
```{r}
make.names(names(test_df))
```

This function is powered by the underlying exported function **`make_clean_names()`**, which accepts and returns a character vector of names (see below).  This allows for cleaning the names of *any* object, not just a data.frame.  `clean_names()` is retained for its convenience in piped workflows, and can be called on an `sf` simple features object or a `tbl_graph` tidygraph object in addition to a data.frame.   

### Do those data.frames actually contain the same columns?
#### Check with `compare_df_cols()`

For cases when you are given a set of data files that *should* be identical, and you wish to read and combine them for analysis.  But then `dplyr::bind_rows()` or `rbind()` fails, because of different columns or because the column classes don't match across data.frames.

`compare_df_cols()` takes unquoted names of data.frames / tibbles, or a list of data.frames, and returns a summary of how they compare.  See what the column types are, which are missing or present in the different inputs, and how column types differ.

```{r}
df1 <- data.frame(a = 1:2, b = c("big", "small"))
df2 <- data.frame(a = 10:12, b = c("medium", "small", "big"), c = 0, stringsAsFactors = TRUE) # here, column b is a factor
df3 <- df1 %>%
  dplyr::mutate(b = as.character(b))

compare_df_cols(df1, df2, df3)

compare_df_cols(df1, df2, df3, return = "mismatch")
compare_df_cols(df1, df2, df3, return = "mismatch", bind_method = "rbind") # default is dplyr::bind_rows
```

`compare_df_cols_same()` returns `TRUE` or `FALSE` indicating if the data.frames can be successfully row-bound with the given binding method:
```{r}
compare_df_cols_same(df1, df3)
compare_df_cols_same(df2, df3)
```

## Exploring

### `tabyl()` - a better version of `table()`
`tabyl()` is a tidyverse-oriented replacement for `table()`.  It counts combinations of one, two, or three variables, and then can be formatted with a suite of `adorn_*` functions to look just how you want.  For instance:

```{r}
mtcars %>%
  tabyl(gear, cyl) %>%
  adorn_totals("col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns() %>%
  adorn_title()
```

Learn more in the [tabyls vignette](https://sfirke.github.io/janitor/articles/tabyls.html).

### Explore records with duplicated values for specific combinations of variables with `get_dupes()`
This is for hunting down and examining duplicate records during data cleaning - usually when there shouldn't be any.

For example, in a tidy data.frame you might expect to have a unique ID repeated for each year, but no duplicated pairs of unique ID & year.  Say you want to check for and study any such duplicated records.

`get_dupes()` returns the records (and inserts a count of duplicates) so you can examine the problematic cases:
```{r}
get_dupes(mtcars, wt, cyl) # or mtcars %>% get_dupes(wt, cyl) if you prefer to pipe
```

### Explore relationships between columns with `get_one_to_one()`
This function shows which, if any, columns in a data.frame have one-to-one relationships with each other.

Here is a toy example looking at the first four rows of the *starwars* data.frame from the dplyr package.  The variables are grouped into three sets of one-to-one clusters:
```{r message=FALSE}
library(dplyr)
starwars[1:4,] %>%
  get_one_to_one()
```


# Minor functions
Smaller functions for use in particular situations.  More human-readable than the equivalent code they replace.

## Cleaning

### Manipulate vectors of names with `make_clean_names()`

Like base R's `make.names()`, but with the stylings and case choice of the long-time janitor function `clean_names()`.  While `clean_names()` is still offered for use in  data.frame pipeline with `%>%`, `make_clean_names()` allows for more general usage, e.g., on a vector.

It can also be used as an argument to `.name_repair` in the newest version of `tibble::as_tibble`:
```{r}
tibble::as_tibble(iris, .name_repair = janitor::make_clean_names)
```

### Validate that a column has a `single_value()` per group

This function returns the single value in a column, often used in combination with `dplyr::group_by()` to validate that every value of X has only one associated value of Y.  It will complete that value of Y into missing values in that column.  And if there's more than a single value of Y, its `info` argument can help pinpoint where that occurs.

Take this data.frame.  One pesky value of X has multiple values of Y where it should not:
```{r}
not_one_to_one <- data.frame(
  X = rep(1:3, each = 2),
  Y = c(rep(1:2, each = 2), 1:2))

not_one_to_one

# throws informative error:
try(not_one_to_one %>%
      dplyr::group_by(X) %>%
      dplyr::mutate(
        Z = single_value(Y, info = paste("Calculating Z for group X =", X)))
      )
```


### `remove_empty()` rows and columns

Does what it says.  For cases like cleaning Excel files that contain empty rows and columns after being read into R.
```{r}
q <- data.frame(v1 = c(1, NA, 3),
                v2 = c(NA, NA, NA),
                v3 = c("a", NA, "b"))
q %>%
  remove_empty(c("rows", "cols"))
```
Just a simple wrapper for one-line functions, but it saves a little thinking for both the code writer and the reader.

### `remove_constant()` columns

Drops columns from a data.frame that contain only a single constant value (with an `na.rm` option to control whether NAs should be considered as different values from the constant).

`remove_constant` and `remove_empty` work on matrices as well as data.frames.

```{r}
a <- data.frame(good = 1:3, boring = "the same")
a %>% remove_constant()
```

### Directionally-consistent rounding behavior with `round_half_up()`

R uses "banker's rounding", i.e., halves are rounded to the nearest *even* number.  This function, an exact implementation of https://stackoverflow.com/questions/12688717/round-up-from-5/12688836#12688836, will round all halves up.  Compare:
```{r}
nums <- c(2.5, 3.5)
round(nums)
round_half_up(nums)
```

### Round decimals to precise fractions of a given denominator with `round_to_fraction()`

Say your data should only have values of quarters: 0, 0.25, 0.5, 0.75, 1, etc.  But there are either user-entered bad values like `0.2` or floating-point precision problems like `0.25000000001`.  `round_to_fraction()` will enforce the desired fractional distribution by rounding the values to the nearest value given the specified denominator.

There's also a `digits` argument for optional subsequent rounding.


### Fix dates stored as serial numbers with `excel_numeric_to_date()`
Ever load data from Excel and see a value like `42223` where a date should be?  This function converts those serial numbers to class `Date`, with options for different Excel date encoding systems, preserving fractions of a date as time (in which case the returned value is of class `POSIXlt`), and specifying a time zone.
```{r}
excel_numeric_to_date(41103)
excel_numeric_to_date(41103.01) # ignores decimal places, returns Date object
excel_numeric_to_date(41103.01, include_time = TRUE) # returns POSIXlt object
excel_numeric_to_date(41103.01, date_system = "mac pre-2011")
```

### Convert a mix of date and datetime formats to date

Building on `excel_numeric_to_date()`, the new functions `convert_to_date()` and `convert_to_datetime()` are more robust to a mix of inputs.  Handy when reading many spreadsheets that *should* have the same column formats, but don't.

For instance, here a vector with a date and an Excel datetime sees both values successfully converted to Date class:
```{r}
convert_to_date(c("2020-02-29", "40000.1"))
```

### Elevate column names stored in a data.frame row

If a data.frame has the intended variable names stored in one of its rows, `row_to_names()` will elevate the specified row to become the names of the data.frame and optionally (by default) remove the row in which names were stored and/or the rows above it.

```{r}
dirt <- data.frame(X_1 = c(NA, "ID", 1:3),
           X_2 = c(NA, "Value", 4:6))

row_to_names(dirt, 2)
```

### Find the header row buried within a messy data.frame

The function `find_header()` is a companion function to `row_to_names()`.  By default it will search a data.frame for the first row with no missing values and return that row number.

It can also be used to return the row number where a given string is present in the first column, or in any specific column.  Then this result can be supplied to `row_to_names()`.


## Exploring

### Count factor levels in groups of high, medium, and low with `top_levels()`

Originally designed for use with Likert survey data stored as factors.  Returns a `tbl_df` frequency table with appropriately-named rows, grouped into head/middle/tail groups.

+ Takes a user-specified size for the head/tail groups
+ Automatically calculates a percent column
+ Supports sorting
+ Can show or hide `NA` values.

```{r}
f <- factor(c("strongly agree", "agree", "neutral", "neutral", "disagree", "strongly agree"),
            levels = c("strongly agree", "agree", "neutral", "disagree", "strongly disagree"))
top_levels(f)
top_levels(f, n = 1)
```
