---
title: "tabyls"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{tabyls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Motivation: why tabyl?

It's like `table()`: you count co-occurences of one, two, or three variables.

It's also not like `table()`: it is built for data.frame inputs, returns data.frames, and is friendly to the common formatting choices one expects if coming from Microsoft Excel or SPSS.

## What it is

We could discuss the details, but you probably came here because you just want it to work.  In short, `tabyl()` produces frequency tables using 1, 2, or 3 variables.  Handy, but nothing too special... except that `tabyl()` stores a copy these counts (the "core") as an attribute of the result.  This allows for subsequent `adorn_` functions to add pretty formatting.

## One-way tabyl

```{r one_way}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(janitor)
t1 <- mtcars %>%
  tabyl(carb)

t1
```


When `NA` values are present, `tabyl()` also produces the "valid" percentages, i.e., with missing values removed from the denominator.  And *shhh, a secret:* While `tabyl()` is meant to take a data.frame and column names, you can also get a one-way tabyl by calling it directly on a vector:
```{r one_way_vector}

x <- c("big", "big", "small", "small", "small", NA)
tabyl(x)
```


The `adorn_` helper functions were made for 2-way tabyls, but the ones that would make sense for a 1-way tabyl do work:
```{r one_way_adorns}
t1 %>%
  adorn_totals("row") %>%
  adorn_pct_formatting()
```


## Two-way tabyl

This is often called a "crosstab" (ala Microsoft Access) or "contingency" table.  The basic call just shortens the very common tidyverse pipeline of `dplyr::count`-ing, followed by `tidyr::spread`-ing to wide form:

```{r two_way}
t2 <- mtcars %>%
  tabyl(carb, cyl)

t2
```

But it's the `adorn_` helper functions that make tabyls powerful.  For instance:

```{r two_way_adorns}

t2 %>%
  adorn_totals("col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns()
```

## Three-way tabyl

Just like `table()` can take three variables, so can `tabyl()`, producing a list of tabyls:

```{r three_Way}
t3 <- mtcars %>%
  tabyl(carb, cyl, am)

t3
```

One can use `purrr::map()` to apply the `adorn_` helper functions:

```{r three_way_adorns}
library(purrr)
t3 %>%
  map(adorn_totals, "row") %>%
  map(adorn_percentages, "row") %>%
  map(adorn_pct_formatting, digits = 1) %>%
  map(adorn_ns)
```

## Features of tabyls

+ When called on a factor, it will include missing levels in the result (levels not present in the vector)


## The `adorn_*` functions

These small, modular functions build on a `crosstab()` to  approximate the functionality of a quick Microsoft Excel PivotTable.  They print an elegant result, either for interactive analysis or for sharing in a report, e.g., with `knitr::kable()`.  For example:

```{r}
mtcars %>%
  crosstab(cyl, gear) %>%
  adorn_totals("row") %>%
  adorn_percentages("row") %>% 
  adorn_pct_formatting() %>%
  adorn_ns()

```

The display options can be fine-tuned:

+ Add totals row, column, or both
+ Calculate percentages along either axis or over the entire tabyl
+ Control how many percentage digits to display and whether to append a % sign
+ Round numbers either with the default `round()` function, or round-half-to-up using a [custom rounding function](http://stackoverflow.com/a/12688836/4470365)
    + e.g., round 10.5 up to 11, consistent with Excel's tie-breaking behavior
    + This contrasts with rounding 10.5 down to 10 as in base R's `round(10.5)`.
    
You can also call these functions on other data.frames, not just the result of calls to `crosstab()`.  E.g., `mtcars %>% adorn_totals("col") %>% adorn_percentages("col")`.

** DISCUSS THE ORDER THEY GET CALLED IN **

## BYOt (Bring Your Own tabyl)

Use `as_tabyl()` when you want to apply the `adorn_` functions to something that is not the result of a call to `tabyl()`.  For instance:

**TODO: do you actually need to call as_tabyl, ever?  Or is it applied automatically by `adorn_` calls?**
```{r as_tabyl}
library(dplyr)
percent_above_20_mpg <- mtcars %>%
  group_by(cyl) %>%
  summarise(pct_above_20 = mean(mpg > 20))

percent_above_20_mpg %>%
  adorn_pct_formatting()
```