---
title: "tabyls: a tidy, fully-featured approach to counting things"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
vignette: >
  %\VignetteIndexEntry{tabyls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk_options, include = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

### Installing

The features of `tabyl()` shown here are in the development version of janitor on GitHub and are not on CRAN yet.  You can install the dev version with `devtools::install_github("sfirke/janitor")`.

## Motivation: why tabyl?

Analysts do a lot of counting.  Indeed, it's been said that "[data science is mostly counting things](https://twitter.com/joelgrus/status/833691273873600512)." But the base R function for counting, `table()`, leaves much to be desired:

- It doesn't accept data.frame inputs (and thus doesn't play nicely with the tidyverse)
- It doesn't output data.frames
- Its results are hard to format.  Compare the look and formatting choices of an R table to a Microsoft Excel PivotTable or even the table formatting provided by SPSS.

`tabyl()` is an approach to tabulating variables that addresses these shortcomings.  It's part of the janitor package because counting is such a fundamental part of data cleaning and exploration.

`tabyl()` is tidyverse-aligned and is primarily built upon the dplyr and tidyr packages.

## How it works

On its surface, `tabyl()` produces frequency tables using 1, 2, or 3 variables.  Under the hood, `tabyl()` also attaches a copy of these counts as an attribute of the resulting data.frame.

The result looks like a basic data.frame of counts, but because it's also a `tabyl` containing this metadata, you can use `adorn_` functions to add pretty formatting.

# Examples
This vignette demonstrates `tabyl` in the context of studying humans in the `starwars` dataset from dplyr:
```{r clean_starwars, warning = FALSE, message = FALSE}
library(dplyr)
humans <- starwars %>%
  filter(species == "Human")
```

## One-way tabyl

Tabulating a single variable is the simplest kind of tabyl:

```{r one_way}
library(janitor)

t1 <- humans %>%
  tabyl(eye_color)

t1
```


When `NA` values are present, `tabyl()` also displays "valid" percentages, i.e., with missing values removed from the denominator.  And while `tabyl()` is built to take a data.frame and column names, you can also produce a one-way tabyl by calling it directly on a vector:

```{r one_way_vector}
x <- c("big", "big", "small", "small", "small", NA)
tabyl(x)
```


Most `adorn_` helper functions are built for 2-way tabyls, but those that make sense for a 1-way tabyl do work:
```{r one_way_adorns}
t1 %>%
  adorn_totals("row") %>%
  adorn_pct_formatting()
```


## Two-way tabyl

This is often called a "crosstab" or "contingency" table.  The initial call produces the same result as the common combination of `dplyr::count()`, followed by `tidyr::spread()` to wide form:

```{r two_way}
t2 <- humans %>%
  tabyl(gender, eye_color)

t2
```

And since it's a `tabyl`, we can enhance it with `adorn_` helper functions.  For instance:

```{r two_way_adorns}

t2 %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns()
```

Adornments have options to control axes, rounding, and other relevant formatting choices (more on that below).

## Three-way tabyl

Just as `table()` accepts three variables, so does `tabyl()`, producing a list of tabyls:

```{r three_Way}
t3 <- humans %>%
  tabyl(eye_color, skin_color, gender)

t3 # the result is a tabyl of eye color x skin color, split into a list by gender
```

Use `purrr::map()` to apply the `adorn_` helper functions to the entire list:

```{r three_way_adorns, warning = FALSE, message = FALSE}
library(purrr)
humans %>%
  tabyl(eye_color, skin_color, gender, show_missing_levels = FALSE) %>%
  map(adorn_totals, "row") %>%
  map(adorn_percentages, "all") %>%
  map(adorn_pct_formatting, digits = 1) %>%
  map(adorn_ns)

```

### Other features of tabyls

+ When called on a factor, it will include missing levels (levels not present in the data) in the result
    + This can be suppressed if not desired
+ `NA` values can be displayed or suppressed
+ Prints without displaying row numbers

## `adorn_*` functions

These modular functions build on a `tabyl` to  approximate the functionality of a quick PivotTable in Microsoft Excel.  They print elegant results for interactive analysis or for sharing in a report, e.g., with `knitr::kable()`.  For example:

```{r}
humans %>%
  tabyl(gender, eye_color) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("row") %>% 
  adorn_pct_formatting(rounding = "half up", digits = 0) %>%
  adorn_ns() %>%
  knitr::kable()

```

### The adorn functions are:

+ **`adorn_totals_row()`**: Add totals row, column, or both.  Replaces the janitor functions `add_totals_row` and `add_totals_col`
+ **`adorn_percentages()`**: Calculate percentages along either axis or over the entire tabyl
+ **`adorn_pct_formatting()`**: Format percentage columns, controlling number of digits to display and whether to append the `%` symbol
+ **`adorn_rounding()`**: Round a data.frame of numbers (usually the result of `adorn_percentages`), either using the base R `round()` function or rounding all ties up using a custom rounding function ([thanks, StackOverflow](http://stackoverflow.com/a/12688836/4470365)).
    + e.g., round 10.5 up to 11, consistent with Excel's tie-breaking behavior.
      + This contrasts with rounding 10.5 down to 10 as in base R's `round(10.5)`.
    + `adorn_rounding()` outputs retain the class `numeric`, allowing for graphing, sorting, etc.  It's a less-aggressive substitute for `adorn_pct_formatting()`; these two functions should not be called together.
+ **`adorn_ns()`**: add Ns to a tabyl.  These can be drawn from the tabyl's `core` attribute (by default), or supplied by the user.


These adornments should be called in a logical order, e.g., you probably want to add totals before percentages are calculated.  In general, call them in the order they appear above.

Users of janitor version <= 0.3.0 should replace the obsolete `adorn_crosstab()` with combinations of the above `adorn_` functions.

## BYOt (Bring Your Own tabyl)

You can also call `adorn_` functions on other data.frames, not only the results of calls to `tabyl()`.  E.g., `mtcars %>% adorn_totals("col") %>% adorn_percentages("col")` performs as expected, despite `mtcars` not being a `tabyl`.

This can be handy when you have a data.frame that is not a simple tabulation generated by `tabyl` but would still benefit from the `adorn_` formatting functions.

A simple example: formatting percentages in a data.frame showing the % of records meeting a certain condition:

```{r first_non_tabyl}
percent_above_165_cm <- humans %>%
  group_by(gender) %>%
  summarise(pct_above_165_cm = mean(height > 165, na.rm = TRUE))

percent_above_165_cm %>%
  adorn_pct_formatting()
```

Here's a more complex example.  We'll create a table containing the mean of a 3rd variable when grouped by two other variables, then use `adorn_` functions to round the values and append Ns.  The first part is pretty straightforward: 
```{r more_non_tabyls, warning = FALSE, message = FALSE}
library(tidyr) # for spread()
mpg_by_cyl_and_am <- mtcars %>%
  group_by(cyl, am) %>%
  summarise(mpg = mean(mpg)) %>%
  spread(am, mpg)

mpg_by_cyl_and_am
```

Now to `adorn_` it.  Since this is not a result of a `tabyl()` call, it doesn't have the underlying Ns stored in the `core` attribute, so we'll have to supply them:
```{r add_the_Ns}
mpg_by_cyl_and_am %>%
  adorn_rounding() %>%
  adorn_ns(
    ns = mtcars %>% # calculate the Ns on the fly by calling tabyl on the original data
      tabyl(cyl, am)
  )
```


Or you could tinker with the Ns before appending them, e.g., if you have large Ns in a tabyl, divide them by 1000, round, and append "k" before calling `adorn_ns`.

### Questions?  Comments?

File [an issue on GitHub](https://github.com/sfirke/janitor/issues) if you have questions or ideas related to `tabyl()` and its `adorn_` helpers or encounter problems while using them.